---
title: "California Proposition 99"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{California Proposition 99}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(width = 88)
```

This vignette walks through the classic Abadie–Diamond–Hainmueller (2010)
synthetic control for California's Tobacco Control Program (Proposition 99).
A Chinese teaching version is in `vignettes/prop99-zh.Rmd`.

## The idea

We want the effect of Proposition 99 on California cigarette sales, but there
is no randomized control state. Synthetic control builds a *weighted average
of donor states* that tracks California on pre-treatment predictors, then
reads the post-1989 gap as the treatment effect.

Donor weights \(W\) are nonnegative and sum to one. Predictor weights \(V\)
are a diagonal matrix that says how much each predictor counts in that match.
`synthaio` estimates the same Abadie SCM as Stata `synth` / `synth2` and
returns a synth2-class report (weights, balance, optional placebos and
leave-one-out). It is not gsynth, not synthetic DID, and not a wrapper around
`tidysynth`.

California is unit `3` in the bundled `smoking` panel. Proposition 99 passed
in November 1988; the usual treatment date in this specification is 1989.

## Estimate

The gold specification matches the smoking / `synth2, nested allopt` fixture:
average the bare covariates over 1980–1988, and also match cigarette sales in
1988, 1980, and 1975. `allopt = TRUE` runs multi-start nested \(V\)
optimization.

```{r fit}
library(synthaio)
data("smoking", package = "synthaio")

fit <- scm(
  cigsale ~ lnincome + age15to24 + retprice + beer +
    cigsale(1988) + cigsale(1980) + cigsale(1975),
  data = smoking, unit = "state", time = "year",
  treat = 3, trperiod = 1989, xperiod = 1980:1988,
  allopt = TRUE
)
fit
```

```{r helpers, include = FALSE}
state_lab <- c(
  "3" = "California",
  "4" = "Colorado",
  "5" = "Connecticut",
  "19" = "Montana",
  "21" = "Nevada",
  "34" = "Utah"
)
w <- coef(fit)
nz <- sort(w[w > 0], decreasing = TRUE)
nz_labs <- unname(state_lab[names(nz)])
v_ord <- sort(v_weights(fit), decreasing = TRUE)
gap2000 <- unname(effects(fit)[["2000"]])
```

The treated unit is California (`r fit$spec$treat`), treatment starts in
`r fit$spec$trperiod`, and the method is `r fit$method`. Pre-treatment RMSE
is `r sprintf("%.3f", fit$rmse)` packs, \(R^2\) is
`r sprintf("%.3f", fit$r2)`, and the post-1989 average gap (ATT) is
`r sprintf("%.2f", fit$att)` packs per capita. In 2000 the gap is
`r sprintf("%.1f", gap2000)` packs.

## Balance and donor weights

Predictor balance compares California, the synthetic, and the unweighted
donor average. Percent bias is relative to the treated mean.

```{r balance}
balance(fit)
```

Nonzero donor weights (state ids in `smoking`; California is 3):

```{r weights}
nz <- sort(coef(fit)[coef(fit) > 0], decreasing = TRUE)
data.frame(
  state_id = names(nz),
  state = unname(c(
    "4" = "Colorado", "5" = "Connecticut", "19" = "Montana",
    "21" = "Nevada", "34" = "Utah"
  )[names(nz)]),
  weight = as.numeric(nz),
  row.names = NULL
)
```

This fit uses `r length(nz)` donors:
`r paste(sprintf("%s (%.3f)", nz_labs, as.numeric(nz)), collapse = ", ")`.
That donor *set* is the same five states recovered by Stata
`synth2, nested allopt` (Utah, Nevada, Montana, Colorado, Connecticut).

Predictor weights \(V\) need not match Stata digit-for-digit. This run puts
the most mass on `r names(v_ord)[1]` (`r sprintf("%.3f", v_ord[[1]])`) and
`r names(v_ord)[2]` (`r sprintf("%.3f", v_ord[[2]])`). Published
`synth2, nested allopt` replicas typically concentrate on `age15to24` and
`cigsale(1975)` — a different local minimum of the same nested MSPE surface.
Do not treat an exact \(V\) match as a correctness check; the locked donor
set and RMSE / ATT tolerances live in
`tests/testthat/test-gold-standard.R`.

## Paths and gaps

```{r path}
ggplot2::autoplot(fit, type = "path")
```

```{r effect}
ggplot2::autoplot(fit, type = "effect")
```

The path plot is California versus the synthetic; the effect plot is the
year-by-year gap. The dashed line is 1989.

## Placebos (not run here)

A full in-space placebo refits the specification once for every donor as if
it had been treated. That is 39 nested solves and is too slow for a vignette.
The checked run is `tests/testthat/test-gold-placebo.R` (nested, not allopt).
The code below is the API; it is not evaluated.

```{r placebo, eval = FALSE}
# In-space placebo: every donor as a fake treated unit.
# Prefer nested (not allopt) here; see test-gold-placebo.R.
fit_space <- scm(
  cigsale ~ lnincome + age15to24 + retprice + beer +
    cigsale(1988) + cigsale(1980) + cigsale(1975),
  data = smoking, unit = "state", time = "year",
  treat = 3, trperiod = 1989, xperiod = 1980:1988,
  nested = TRUE,
  placebo = list(unit = TRUE)
)
summary(fit_space)
ggplot2::autoplot(fit_space, type = "gaps")
ggplot2::autoplot(fit_space, type = "mspe")

# In-time placebo: a fake treatment date before 1989.
fit_time <- scm(
  cigsale ~ lnincome + age15to24 + retprice + beer +
    cigsale(1988) + cigsale(1980) + cigsale(1975),
  data = smoking, unit = "state", time = "year",
  treat = 3, trperiod = 1989, xperiod = 1980:1988,
  allopt = TRUE,
  placebo = list(period = 1985)
)

# Leave-one-out: drop each nonzero donor in turn.
fit_loo <- scm(
  cigsale ~ lnincome + age15to24 + retprice + beer +
    cigsale(1988) + cigsale(1980) + cigsale(1975),
  data = smoking, unit = "state", time = "year",
  treat = 3, trperiod = 1989, xperiod = 1980:1988,
  allopt = TRUE,
  loo = TRUE
)
ggplot2::autoplot(fit_loo, type = "loo")
```

## References

Abadie, A., Diamond, A., and Hainmueller, J. (2010). Synthetic Control
Methods for Comparative Case Studies: Estimating the Effect of California's
Tobacco Control Program. *Journal of the American Statistical Association*,
105(490), 493–505.

Yan, G., and Chen, Q. (2023). synth2: Synthetic control method with placebo
tests, robustness test, and visualization. *The Stata Journal*, 23(3),
597–624.
